---
title: "STAT302Package Tutorial"
author: "Brad Scurlock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STAT302Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--- Begin styling code. --->
<style type="text/css">
/* Whole document: */
/*
body{
  font-family: "Palatino Linotype", "Book Antiqua", Palatino, serif;
  font-size: 12pt;
}
*/
h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author {
  font-size: 18px;
  text-align: center;
}
</style>
<!--- End styling code. --->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## **Introduction**

This is Project 3, the final project STAT 302 (Statistical Computing) at the University of Washington.

Install from GitHub using:

```{r, eval = FALSE}
devtools::install_github("smokingcrater/STAT302Package")
```

Load the package.

```{r setup}
library(STAT302Package)
```

## **my_t.test()**

my_t.test() is a blah, blah, blah

Mention using gapminder data and life expectancy

```{r}
summary(my_gapminder$lifeExp)
```

In the following three hypothesis tests, the significance level (alpha) for p-value cut-off is 0.05.

```{r}
my_t.test(my_gapminder$lifeExp, alternative = "two.sided", mu = 60)
```

The above p-value is nominally greater than alpha, thus indicating weak evidence against the null hypothesis.  Therefore we fail to reject the null hypothesis.

```{r}
my_t.test(my_gapminder$lifeExp, alternative = "less", mu = 60)
```

The above p-value is slightly less than alpha, thus indicating strong evidence against the null hypothesis.  Therefore we reject the null hypothesis.

```{r}
my_t.test(my_gapminder$lifeExp, alternative = "greater", mu = 60)
```

The above p-value is significantly greater than alpha, thus indicating weak evidence against the null hypothesis.  Therefore we fail to reject the null hypothesis.

## **my_lm()**

my_lm() is a blah, blah, blah

```{r}
my_lm(formula = lifeExp ~ gdpPercap + continent, data = my_gapminder)
```

Linear regression was performed on the gapminder data set using lifeExp as the response variable (y) and gdpPercap and continent as the exploratory variables (x)

Carefully interpret the gdpPercap coefficient blah, blah, blah

```{r}
summary(my_gapminder$gdpPercap)
```

In the following hypothesis test, the significance level (alpha) for p-value cut-off is 0.05.

```{r}
my_t.test(my_gapminder$gdpPercap, alternative = "two.sided", mu = 7215.3)
```

The above p-value is substantially greater than alpha, thus indicating weak evidence against the null hypothesis.  Therefore we fail to reject the null hypothesis.

```{r}
# To obtain fitted values, multiply X %*% beta where:
#   X is the result of model.matrix()
#   beta represents the estimates returned from my_ln().
fitted_life_exp <-
  model.matrix(lifeExp ~ gdpPercap + continent, data = my_gapminder) %*%
  my_lm(formula = lifeExp ~ gdpPercap + continent, data = my_gapminder)[, 1]

# Create a data frame of the actual and fitted values along with their
# associated continent.
actual_vs_fit_df <-
  data.frame(actual = my_gapminder$lifeExp,
             fitted = fitted_life_exp,
             continent = my_gapminder$continent)

# Plot actual vs fitted values and use color to distinguish continents.
ggplot2::ggplot(actual_vs_fit_df,
                ggplot2::aes(x = fitted,
                             y = actual,
                             col = continent)) +
  ggplot2::geom_point(size = 0.75) +
  ggplot2::geom_abline(slope = 1, intercept = 0, col = "red", lty = 2) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::scale_color_manual(name = "Continent",
                             values = c("Africa" = "orange",
                                        "Americas" = "yellow",
                                        "Asia" = "green",
                                        "Europe" = "blue",
                                        "Oceania" = "pink")) +
  ggplot2::labs(x = "Fitted values",
                y = "Actual values",
                title = "Actual versus Fitted") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5,
                                                    face = "bold"))
```

## **my_knn_cv()**

my_knn_cv() is a blah, blah, blah

```{r}
```

## **my_rf_cv()**

my_rf_cv() is a blah, blah, blah

```{r}
```

## **Conclusion**

In summary, blah, blah, blah
